%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% abstrac-en.tex
%% NOVA thesis document file
%%
%% Abstract in English
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Currently, offloading storage and processing capacity to cloud servers is a growing
trend among web-enabled services managing big datasets. This happens because high
storage capacity and powerful processors are expensive, whilst cloud services provide
cheaper, ongoing, elastic, and reliable solutions. The problem with this cloud-based outsourced solutions are that they are highly accessible through the Internet, which is good,
but therefore can be considerably exposed to attacks, out of users’ control. By exploring
subtle vulnerabilities present in cloud-enabled applications, management functions, operating systems and hypervisors, an attacker may compromise the supported systems,
thus compromising the privacy of sensitive user data hosted and managed in it. These
attacks can be motivated by malicious purposes such as espionage, blackmail, identity
theft, or harassment. A solution to this problem is processing data without exposing it to
untrusted components, such as vulnerable OS components, which might be compromised
by an attacker.

In this thesis, we intend to study existent technologies capable of enabling applica-
tions to run on top of trusted environments and conduct an experimental evaluation to
adopt such approaches to our solution, in order to help to deploy existing applications
on top of Intel-SGX, with overheads comparable to applications modified to use other
trustable security layers. Our targeted evaluation will be conducted in designing TREDIS
- a Trusted Full-Fledged REDIS Key-Value Store solution, redesigning and implementing it as a full-fledged solution to be offered as a Trusted Cloud-enabled Platform as
a Service. This includes the possibility to support a secure REDIS-cluster architecture
supported by docker-virtualized services running in independent SGX-enabled instances,
with operations running on always-encrypted in-memory datasets.

% Palavras-chave do resumo em Inglês
\begin{keywords}
Intel SGX, REDIS, Trusted Computing, Trusted Execution Environments, Data Protection, Privacy-Preservation, Dependable In-Memory Key-Value Stores
\end{keywords} 


